Starcode: an exact algorithm for sequence clustering

### 2014-07-03 doc/document.tex
Added counts for N. meningitidis.
Written 'scalability' section. Moved 'Benchmark options' to Results section. Corrected some typos.

\begin{abstract}
\section{Motivation:}
The increasing throughput of sequencing technologies offers new applications
and challenges for computational biology. One such application is the use of
random barcodes to trace and quantify transcripts or lineages in experimentals
setups. The high error rate of modern sequencers calls for additional
post-processing techniques capable of detecting and reverting the misreads.
However, in the absence of a reference population, the problem amounts to
performing a pairwise comparison of all the barcodes, which is unfeasible for
excessive computationally complexity.

@@

\subsection{Lookup search}
If the Levenshtein distance between two sequences of length $L$ is
less then or equal to $\tau$, then there must exist a perfect
match of length at least $k = L / (\tau+1)$, assuming for the time being
that $k$ is an integer.
We take advantage of this property to build a lookup table containing
all such $k$-mers from sequences inserted in the trie. When
a new sequence is queried, subsequences are first extracted from it
and queried against the lookup table. If none of the subsequences
are found, the minimum distance between the query and all the
sequences in the trie is at least $\tau+1$, and the trie search can
therefore be omitted. This lookup is particulary useful for long query
sequences (see \ref{scalability}). In this case $k$ is large, so the
probability of an exact match between subsequences of length $k$ is small.
The absence of match is detected at the computational cost of several
table lookups, compared to the higher cost of measuring the Levenshtein
distance between long sequences.

In general $L / (\tau+1)$ is not an integer. To deal
with this complication, we divide the query into $\tau+2$ subsequences.
The first $k+1$ are called words and have target length
$(L-\tau) / (\tau+1)$. The last subsequence is called the tail and has
length $\tau$ for reasons explained below. The length of the words is
computed by integer division, and the remainder $R$ is distributed
equally among the last $R$ words. As a result, the words have a length
that differs by at most 1, with a total length equal to $L-\tau$.

A word at the beginning of the query is not a match for the same
word at the end of a sequence in the trie. For this reason, we query
the words against $k+1$ distinct lookup tables. In order to allow
insertions and deletions, the lookup search must also query shifted
versions of the words. In the worst case, only one word is found and the
mismatches are evenly distributed in the query sequence. The maximum
number of cumulative insertions/deletions at the $i$-th word
is $i-1$. Hence, the $i$-th word has to be shifted and queried against
the lookup table $2(i-1)+1$ times to cover all the possible cases.
The last word will therefore be queried $2\tau+1$ times, including
$\tau$ shifts on the right and $\tau$ shifts on the left. For the latter
we use the $\tau$ nucleotides stored in the tail.

@@

\begin{algorithm}
  \caption{Starcode algorithm}
  \label{alg:starcode}
  \begin{algorithmic}[1]
    \State \textbf{Define:} $\tau$
    \State \textbf{Variables:} $seed$, $start = 0$,
    $height$, $seq$, $trie$, $lastseq$, $k$
    \State \textbf{Containers:} $hits$, $pebbles$
    \State \textsc{read} sequence file
    \State $height \gets$ \textsc{determine} maximum sequence length
    \State \textsc{pad} sequences up to $height$
    \State \textsc{sort} sequences alphabetically
    \State $k \gets$ \textsc{compute} lookup word lengths
    \State $trie \gets$ \textsc{create} an empty trie of height $height$
    \State \textsc{insert} root node of $trie$ in $pebbles$ at depth 0
    \ForAll{sequences}
    \State $seq \gets$ \textsc{get} next sequence
    \If{at least one $k$-mer of $seq$ is in the lookup table}
    \State $seed \gets$ \textsc{length} of shared prefix
    between current and next sequence
    \State $start \gets$ \textsc{length} of shared prefix
    between $seq$ and $lastseq$
    \State \textsc{clear} hits
    \State \textsc{clear} $pebbles$ at depth $>start$
    \ForAll{$pebbles$ at depth $start$}
    \State $node \gets$ \textsc{get} next node from $pebbles$
    \State \textbf{call} \textsc{poucet}($seq$, $node$, $seed$, $hits$)
    \EndFor
    \State \textsc{process} $hits$ and \textsc{link} matches to $seq$
    \State $lastseq \gets seq$
    \EndIf
    \State \textsc{insert} $seq$ path in $trie$
    \State \textsc{insert} $seq$ $k$-mers into the lookup table
    \EndFor
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}
  \caption{Poucet search algorithm}
  \label{alg:poucet}
  \begin{algorithmic}[1]
    \Procedure{poucet}{$query$, $node$, $seed$, $hits$}:
    \State \textsc{compute} $node$-specific column values following NW
    algorithm \Comment{Fig.1}
    \ForAll{$child$ nodes in $node$}
    \State \textsc{compute} $child$-specific row values following NW
    algorithm \Comment{Fig.1}
    \State \textsc{compute} center value using row and column values \Comment{Fig.1}
    \If{center value $>\tau$} \Comment{Mismatches exceeded.}
    \State \textbf{continue} with next $child$
    \EndIf
    \If{$node$ depth = $height$} \Comment{Hit found.}
    \State \textsc{save} $node$ sequence in $hits$
    \State \textbf{continue} with next $child$
    \EndIf
    \If{$node$ depth $\leq seed$}
    \State \textsc{save} $node$ in pebbles at current depth
    \EndIf
    \State \textbf{call} poucet($child$, $seed$, $hits$)
    \EndFor
    \EndProcedure
  \end{algorithmic}
\end{algorithm}

\subsection{Parallelization}
To parallelize the search, queries are separated into contiguous blocks after
sorting. The matching algorithm proceeds in two phases. During the first phase,
an independent trie is created and filled with the sequences
of one block using the seek and construct process described above. In the second
phase, blocks of sequences are queried against but not inserted into the tries
built in the first phase. If the
queries are segregated into $N$ blocks, the first phase consists of $N$ seek and
construct jobs, and the second consists of $N(N-1)/2$ query jobs. Since the jobs
show little dependence on each other, the matching algorithm can be efficiently
parallelized provided $N$ is larger than the number of independent threads.

\subsection{Clustering}
Starcode implements a multi-purpose clustering algorithm called ``sphere
clustering'', and a message passing algorithm \citep{mackay} tailored for the
task of clustering random barcodes. In sphere clustering, barcodes are sorted
by frequency of occurrence and each barcode, starting from the most frequent,
can claim its $d$-neighbors that were not already claimed. This is the method
used by \cite{pmid23953119}.

In message passing clustering, read counts are distributed equally among the
closest neighbors of each barcode if they are at least 5 times more frequent.
The remaining sequences at the end of the process are considered
canonical, and their associated count is the estimated cluster size. A barcode
is assigned to a cluster if all its read counts are eventually given to the
corresponding canonical barcode. The barcodes for which the read counts are
split between different canonical barcodes are not assigned to any cluster. The
reason for imposing a factor 5 or larger in order to transfer the read counts
is that barcodes with similar frequencies are not likely derived from each other
through sequencing errors. More likely they are either unrelated, or they both
derive from another more abundant barcode.

\subsection{Benchmark options}

The benchmark dataset is described in the Results section. All software were
set to run with 20 parallel threads on a 20-core Intel Xeon processor. The
execution options for each software are summarized in
Table~\ref{tab:soft-options}.

Denoted as \emph{sequential R} is the sequential algorithm of
\cite{pmid23953119} which is implemented in R and relied on the \texttt{stringdist}
function of the \texttt{stringdist} package.

@@

\section{Results}
\subsection{Scalability}
\label{scalability}
Write some crap here.

\begin{figure}[!tpb]%figurex
\centerline{\includegraphics[scale=0.47]{scalability.pdf}}
\caption{Scalability. Top left: the running time of starcode is
approximlately $O(n^{1.7})$, where $n$ is the number of sequences to
cluster. Top right: the running time is very sensitive to the maximum
Levenshtein distance used in custering.  Bottom left: the running time
as a function of the length of the sequences has a complex behavior. It
increases initially but then decreases for long sequences because the
lookup search becomes more effective. Bottom right: the scaling of the
relative performance is sub-linear as the number of threads increases.
The algorithm is fully parallel, but the memory bandwidth becomes
limiting on the tested hardware.
}\label{fig:X}
\end{figure}

@@

Using 12 cores, starcode clustered the shuffled 50 million sequences
in 5 minutes and 17 seconds without a single error (the output
consisted of 1 million clusters of size 50).  The running times
of CD-HIT and USEARCH were [59] and [64] minutes respectively, and they
identified 2,362,352 and 2,739,954 clusters respectively. The results
are summarized in Figure~\ref{fig:benchmark}. Both algorithms rely on
heuristics, explaining why the number of clusters is off target. These
results clearly demonstrate that starcode outperforms other clustering
algorithms in speed and accuracy to cluster random sequences.

>>>>

\begin{abstract}
\section{Motivation:}
The increasing throughput of sequencing technologies offers new applications
and challenges for computational biology. One such application is the use of
random barcodes to trace and quantify transcripts or lineages in experimentals
setups. The high error rate of modern sequencers calls for additional
post-processing techniques capable of detecting and reverting the misreads.
However, in the absence of a reference population, the problem amounts to
performing a pairwise comparison of all the barcodes, which is unfeasible for
excessive computational complexity.

@@

\subsection{Lookup search}
If the Levenshtein distance of two sequences of length $L$ is
less than or equal to $\tau$, there must exist a perfect match between
them of at least $k =\lfloor L / (\tau+1)\rfloor$ nucleotides. We take
advantage of this property to build a lookup table containing all such
$k$-mers of the sequences already present in the trie. When a new
sequence is queried, all its subsequences are extracted and queried
against the lookup table. If none of the subsequences are found, the
minimum distance between the query and all the sequences in the trie
is necessarily greater than $\tau$, and the trie search can therefore
be omitted. This lookup is particulary useful for long query sequences
(see \ref{scalability}), where $k$ is large and the probability
of an exact match between subsequences is small unless they are
actually related, i.e. they belong to the same cluster. With this
method, the absence of match is detected at the computational cost of
several table lookups, compared to the higher cost of measuring the
Levenshtein distance between long sequences.

In general, however, $L / (\tau+1)$ is not an integer. To deal
with this complication, we divide the query into $\tau+2$ subsequences.
The first $k+1$ are called \emph{words} and have a target length
$(L-\tau) / (\tau+1)$. The last subsequence is called \emph{tail} and has
length $\tau$. The length of the words is computed by integer
division, and the remainder $R$ is distributed equally among the last
$R$ words. As a result, the words have a length that differs by at
most 1, with an aggregated length equal to $L-\tau$.

Since two exact words at different positions in the sequence shall not
represent a lookup match, we store and query the words against $k+1$
distinct lookup tables. In order to allow insertions and deletions, 
the lookup search must also query shifted versions of each word. In
the worst matching case, only one word is found and the mismatches are
evenly distributed in the query sequence. In such case, the maximum 
number of cumulative insertions/deletions at the $i$-th word is
$i-1$. Hence, the $i$-th word has to be shifted and queried against
the lookup table $2(i-1)+1$ times to cover all the possible
cases. Hence, the last word will be queried $2\tau+1$ times, including
$\tau$ shifts on the right and $\tau$ shifts on the left. For the
latter shift we make use of the $\tau$ nucleotides stored in the
tail. The total number of lookups per sequence for a given $\tau$ is
$(\tau+1)^2$.

@@

\begin{algorithm}
  \caption{Starcode algorithm}
  \label{alg:starcode}
  \begin{algorithmic}[1]
    \State \textbf{Define:} $\tau$
    \State \textbf{Variables:} $seed$, $start = 0$,
    $height$, $seq$, $trie$, $lastseq$, $k$
    \State \textbf{Containers:} $hits$, $pebbles$
    \State \textsc{read} sequence file
    \State $height \gets$ \textsc{determine} maximum sequence length
    \State \textsc{pad} sequences up to $height$
    \State \textsc{sort} sequences alphabetically
    \State $k \gets$ \textsc{compute} lookup word lengths
    \State $trie \gets$ \textsc{create} an empty trie of height $height$
    \State \textsc{insert} root node of $trie$ in $pebbles$ at depth 0
    \ForAll{sequences}
    \State $seq \gets$ \textsc{get} next sequence
    \If{at least one $k$-mer of $seq$ is in the lookup table}
    \State $seed \gets$ \textsc{length} of shared prefix
    between current and next sequence
    \State $start \gets$ \textsc{length} of shared prefix
    between $seq$ and $lastseq$
    \State \textsc{clear} hits
    \State \textsc{clear} $pebbles$ at depth $>start$
    \ForAll{$pebbles$ at depth $start$}
    \State $node \gets$ \textsc{get} next node from $pebbles$
    \State \textbf{call} \textsc{poucet}($seq$, $node$, $seed$,
    $hits$, $pebbles$)
    \EndFor
    \State \textsc{process} $hits$ and \textsc{link} matches to $seq$
    \State $lastseq \gets seq$
    \EndIf
    \State \textsc{insert} $seq$ path in $trie$
    \State \textsc{insert} $seq$ $k$-mers into the lookup table
    \EndFor
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}
  \caption{Poucet search algorithm}
  \label{alg:poucet}
  \begin{algorithmic}[1]
    \Procedure{poucet}{$query$, $node$, $seed$, $hits$, $pebbles$}:
    \State \textsc{compute} $node$-specific column following NW
    \Comment{Fig.1}
    \ForAll{$child$ nodes in $node$}
    \State \textsc{compute} $child$-specific row following NW
    \Comment{Fig.1}
    \State \textsc{compute} center value using row and column \Comment{Fig.1}
    \If{center value $>\tau$} \Comment{Mismatches exceeded.}
    \State \textbf{continue} with next $child$
    \EndIf
    \If{$node$ depth = $height$} \Comment{Hit found.}
    \State \textsc{save} $node$ sequence in $hits$
    \State \textbf{continue} with next $child$
    \EndIf
    \If{$node$ depth $\leq seed$}
    \State \textsc{save} $node$ in $pebbles$ at current depth
    \EndIf
    \State \textbf{call} poucet($query$, $child$, $seed$, $hits$, $pebbles$)
    \EndFor
    \EndProcedure
  \end{algorithmic}
\end{algorithm}

\subsection{Parallelization}
To parallelize the search, queries are separated into contiguous blocks after
sorting. The matching algorithm proceeds in two phases. During the
first phase, an independent trie is created and filled with the
sequences of its associated block using the seek and construct process
described above. In the second phase, all the other blocks of
sequences are queried against but not inserted into each trie built in
the first phase. If the queries are segregated into $N$ blocks, the
first phase consists of $N$ seek and construct jobs, whereas the second
consists of $N(N-1)/2$ query jobs. Since the jobs show little
dependence on each other, the matching algorithm can be efficiently
parallelized provided $N$ is larger than the number of independent
threads.
\subsection{Clustering}
Starcode implements a multi-purpose clustering algorithm called ``sphere
clustering'' \citep{pmid23953119}, and a message passing algorithm
\citep{mackay} tailored for the task of clustering random barcodes. In
sphere clustering, barcodes are sorted by frequency of occurrence and
each barcode, starting from the most frequent, can claim its
$d$-neighbors that were not already claimed.

In message passing clustering, read counts are distributed equally
among the closest neighbors of each barcode only if they are at least
5 times more frequent. The remaining sequences at the end of the
process are considered canonical, and their associated count is the
estimated cluster size. A barcode is assigned to a cluster if all its
read counts are eventually given to the corresponding canonical
barcode. The barcodes for which the read counts are split between
different canonical barcodes are not assigned to any cluster. The
reason for imposing a factor 5 or larger in order to transfer the read
counts is that barcodes with similar frequencies are not likely
derived from each other through sequencing errors. More likely they
are either unrelated, or they both derive from another more abundant
barcode.

@@

\section{Results}

\subsection{Scalability}
\label{scalability}
In Figure~\ref{fig:X} we present an evaluation of the performance and
scalability of starcode as a function of many critical parameters. The
scalability tests were performed using an **artificial set of random
sequences**. The sequence length and the number of input sequences
varies depending on the test. The standard configuration consists of a
set of **X** sequences of length **Y** running on **Z** threads and a
cluster distance of **D**. To test the scalability of starcode as a
function of a single parameter, only the selected value is modified
whereas the others are kept constant.

In Figure~\ref{fig:X}a we show the running time as a function of the
selected Levenshtein distance for the clusters. The evolution with the
increasing distance shows the common exponential trend. Note, however,
that for distances below 3 the running time remains constant. This is
due to the fact that the average sequence distance of the input
sequences lies above this threshold. Consequently, the constant part
is clearly dependent on the diversity of the input set and when the
average distance between sequences is reached, the running time scales
exponentially.

In Figure~\ref{fig:X}b we measure the running time of starcode as a
function of the number of input sequences $n$. Both axes are in
logarithmic scale, drawing an evident linear evolution. Based on this
result, we conclude that the running time of starcode is approximately
$O(n^{1.7})$.

To prove the effectiveness of the parallel processing, we ran starcode
over the same dataset using different number of threads. As shown in
Figure~\ref{fig:X}c, the algorithm is fully parallel as the relative
performance increases linearly up to 4 threads. The curve bending
produced thereafter is an artifact of having insufficient memory
bandwidth to satisfy the increased demand of memory accesses and it is
due to hardware limitations.

Finally, in Figure~\ref{fig:X}d we investigate the execution time
with varying sequence lengths. It increases initially due to the
ineffectiveness of the lookup table and the limited diversity of short
sequences, i.e. short sequences are less diverse, yielding a bigger
clustering rate. At a certain length, the lookup search algorithm
starts to be effective and boosts the performance since most of the
queries are resolved without the need for trie-search.

\begin{figure}[!tpb]%figurex
\centerline{\includegraphics[scale=0.47]{scalability.pdf}}
\caption{Scalability. a. Running time as a function of the clustering
  distance. b. Logarithm of the running time versus the logarithm of
  the number of sequences to be clustered. c. Relative performance increase for
  different number of parallel threads. d. Running time versus length
  of the input sequences.
}\label{fig:X}
\end{figure}

@@

All software were set to run on a 6-core single-processor Intel Xeon
E5-2620 system with **64~GB** of DDR3-RAM at 1333~Mhz. The execution
options for each software are summarized in
Table~\ref{tab:soft-options}. Denoted as \emph{sequential R} is the
sequential algorithm of \cite{pmid23953119} which is implemented in R
and relies on the \texttt{stringdist} function of the
\texttt{stringdist} package.

Using **12** cores, starcode clustered the shuffled 50 million sequences
in 5 minutes and 17 seconds without a single error (the output
consisted of 1 million clusters of size 50).  The running times
of CD-HIT and USEARCH were [59] and [64] minutes respectively, and they
identified 2,362,352 and 2,739,954 clusters respectively. The results
are summarized in Figure~\ref{fig:benchmark}. Both algorithms rely on
heuristics, explaining why the number of clusters is off target. These
results clearly demonstrate that starcode outperforms other clustering
algorithms both in speed and accuracy with random input sequences.
